Local PCA Shows How Population Structure Differs Along the Genome

### 2016-06-06 writeup/local_pca_paper.tex
peter edited up through methods

\newcommand{\pcone}{PC1}
\newcommand{\pctwo}{PC2}
\newcommand{\given}{\,\vert\,}
\newcommand{\st}{\,\colon\,}
\renewcommand{\and}{\,\&\,}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}

@@

Dimension reduction techniques,
such as principal components,
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship this uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find substantial variation on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is a useful addition to the exploratory toolbox
of population genomics.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{the-method-diagram}
    \end{center}
    \caption{
         The diagram of our method.
         \label{fig:diagram}
    }
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{p{0.8in}p{.8in}p{1in}p{1.2in}p{1.4in}}
  \hline
    species & \# SNPs per window & mean window length (bp) & mean \# windows per chromosome & mean \% variance explained by top 2 PCs \\ 
  \hline
  Human & 100 & 636,494 & 203 & 0.55 \\ 
  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
   \hline
\end{tabular}
\caption{
    Descriptive statistics for each dataset used,
    with the chosen window sizes.
    See text for how window sizes were chosen;
    in each window we performed PCA,
    computed the percent varince explained by the top two principal components;
    the mean across windows is given in the last column.
    \label{tab:data_stats}
}
\end{table}

@@

\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances, since it makes sense conceptually here.}
Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)$ or
$\sum_i (\pcone_{i1} + u_i)$ 
is smaller.

@@

The choice of window length entails a balance between several factors.
In very short windows, genealogies of the samples will only be represented by a few trees,
so variation between windows represents demographic noise rather than meaningful variation in population structure.
Longer windows generally have more distinct trees (and SNPs), 
allowing for less noisy estimation of local population structure.
However, to better resolve meaningful signal, i.e., differences in population structure along the genome, 
we would like reasonably short windows.
Window length choice therefore entails a signal versus noise tradeoff
in the estimates of population structure.
% If we use the first principal component as a measure of population structure, 
% then to choose the best window length,
% we need to find a balance between the standard error of the first principal components for each window 
% and the standard deviation between windows. 
Since we summarize population structure using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
(Recall that the signs for PCs are chosen to match each other.)
Then, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife}:
we divide the $j^\text{th}$ block into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
and we measure noise by averaging over samples and windows:
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

These values are shown calculated separately for each chromosome arm in the \textit{Drosophila} dataset in Table \ref{tab:window_sizes}.
Finally, we choose 100 SNPs, 1000 SNPs and 10000 SNPs as window length for human, \textit{Drosophila}, and \textit{Medicago} respectively.

\begin{table}[ht]
\centering
    \begin{tabular}{rlp{1.1in}rrrrr}
  \hline
 & chrom.\ arm  & window length (in SNPs) & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
  \hline
    & 2L & $\sigma_\text{noise}$  &    4.53 & 4.05 & 3.44 & 1.30 & 0.63 \\ 
    & 	 & $\sigma_\text{signal}$ &    5.25 & 5.19 & 4.72 & 2.60 & 1.77 \\ 
    & 2R & $\sigma_\text{noise}$  &    4.67 & 4.38 & 4.04 & 2.40 & 1.16 \\ 
    & 	 & $\sigma_\text{signal}$ &    5.27 & 5.20 & 5.15 & 4.81 & 4.27 \\ 
    & 3L & $\sigma_\text{noise}$  &    4.56 & 4.47 & 4.05 & 2.71 & 1.57 \\ 
    & 	 & $\sigma_\text{signal}$ &    5.10 & 5.02 & 4.90 & 4.10 & 4.35 \\ 
    & 3R & $\sigma_\text{noise}$  &    4.42 & 4.20 & 3.79 & 2.42 & 1.42 \\ 
    & 	 & $\sigma_\text{signal}$ &    5.08 & 5.01 & 4.94 & 4.43 & 3.74 \\ 
    & X  & $\sigma_\text{noise}$  &    4.98 & 4.52 & 3.92 & 4.02 & 1.30 \\ 
    & 	 & $\sigma_\text{signal}$ &    5.11 & 4.93 & 4.80 & 1.80 & 3.38 \\
   \hline
\end{tabular}
\caption{
    Measures of signal and noise,
    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
    at different window sizes.
    All values are multiplied by $100$
    (so typical variation is of order of 50\% of the actual values).
    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
    starts to be substantially larger than the noise (standard error of PC1 for each window).
    \plr{make 'window length' in bigger bar above the values}
} \label{tab:window_sizes}
\end{table}

\subsection{Similarity of population structure between windows}
We compared population structure in different genomic windows 
using the first two principal components (PCs) and the corresponding eigenvalues from PCA.
We do this, rather than using the entire covariance matrix for computational efficiency 
and because the top PCs summarize important population structure, 
using only these should reduce the effect of noise.
For example, the constructed matrixes for ith and jth window are as following. 
($\lambda _{1i}$ and $\lambda _{2i}$ are the eigenvalues for the first two PCs for \textit{i}th window; 
$M_i$ is the constructed new matrix for \textit{i}th window and \textit{j}th window.
\plr{
Say what $M_i$ and $M_j$ are in the text.  
Also, how about we write $V$ instead of $PC$?  I like to use only one letter for variables, so it's clear it's just one thing, not the product of $P$ and $C$.
And, isn't it $\sqrt{\lambda_{1j}^2+\lambda_{2j}^2}$ on the bottom? Check in the R code (pc{\textunderscore}dist.R).
}

\begin{align}
    M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
\end{align}

The Euclidean distance $D_{ij}$ between the matrices $M_i$ and $M_j$ stands for the similarity of population structure for the \textit{i}th window and \textit{j}th window. 
Due to the orthogonality of eigenvectors,we could use the following method to calculate the pairwise distance greatly saving time and space.

\plr{Define $D$.}

\begin{equation}
    \begin{aligned}
        V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} \qquad
        V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} \\
    \end{aligned}
\end{equation}

\begin{align}
    M_{i} &= V_{i1}V_{i1}^{T}+V_{i2}V_{i2}^{T}
\end{align}

\begin{align}
    \begin{split}
    D_{ij} &= 
        \left \{
                \left ( 
                    V_{i1}\cdot V_{i1}^{} 
                \right )^{2}
                + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
                + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
                +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
        \right. \\ 
        & \left. \qquad {}  
            -2\left [ 
                \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
                + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
                + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
                + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
            \right ]
        \right \}^{1/2}
    \end{split}
\end{align}

Using this procedure, 
we get the pairwise distance matrix that says how similar population structure is 
in each pair of genomic windows.

\subsubsection{Visualize the pairwise distance matrix}
We use Multidimensional scaling (MDS) method to visualize the distance matrices. 
It can reduce the dimensionality of a distance matrix while preserve the distance information between objects as well as possible. 
The result is a set of coordinates for each sample with the property that 
the first M coordinates give the arrangement in M-dimensional space that best recapitulates the original distance matrix.
We use M=2 to produce one or two dimensional visualization of relationships between windows' population structure.

\section{Results}
In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is truly due to population structure, 
since fluctuations due to demographic noise are not expected to show long distance correlations. 
Below, we display the results and investigate likely underlying causes.

>>>>

\newcommand{\pcone}{PC1}
\newcommand{\pctwo}{PC2}
\newcommand{\pc}[1]{PC#1}
\newcommand{\given}{\,\vert\,}
\newcommand{\st}{\,\colon\,}
\renewcommand{\and}{\,\&\,}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\tr}{tr}

@@

Dimension reduction techniques,
such as principal components,
are often used to discover and display large-scale structure in genomic datasets
found in the patterns of kinship
between the genotyped individuals,
and to control for the confounding effects of population structure in genome-wide association studies.
The genome-wide mean kinship this uses
is an average of the relationships across all locus-specific genealogical trees.
However, many biological factors,
including linked selection,
can systematically skew patterns of kinship over intermediate genomic scales.
We show how to use principal components analysis (PCA) to describe this meso-scale variation in kinship,
and apply the method to genomic data from three species.
In each species we find substantial variation on the scale of megabases to tens of megabases.
In a global human dataset, small, discontinuous variation is likely explained by polymorphic chromosomal inversions.
In a dataset of African \textit{Drosophila melanogaster}, large, continuous variation across each chromosome arm
is explained by known chromosomal inversions thought to be under recent selection.
In a range-wide dataset of \textit{Medicago truncatula},
common axes of variation in population structure are shared between chromosomes,
correlate with local gene density,
and may be caused by background selection or local adaptation.
The method is a useful addition to the exploratory toolbox
of population genomics,
and is implemented as an R package.

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{the-method-diagram}
    \end{center}
    \caption{
         An illustration of the method; see text for details.
         % (1) the genome is divided into windows;
         % (2) population structure is summarized in each with PCA, and 
         % (3) dissimilarities for each are computed;
         % (4) the patterns of dissimilarity are visualized with MDS,
         % and (5) genomic regions clustering together are aggregated and visualized, again with PCA.
         \label{fig:diagram}
    }
\end{figure}

@@

\begin{table}[ht]
\centering
    \begin{tabular}{p{0.8in}rrrr}
  \hline
    species 
    & \parbox[t]{.8in}{\# SNPs per \\ window} 
    & \parbox[t]{1in}{mean window\\ length (bp)}
    & \parbox[t]{1.2in}{mean \# windows \\ per chromosome} 
    & \parbox[t]{1.4in}{mean \% variance ex-\\plained by top 2 PCs} \\ 
  \hline
  Human & 100 & 636,494 & 203 & 0.55 \\ 
  \textit{Drosophila melanogaster} & 1,000 & 9,019 & 2,674 & 0.53 \\ 
  \textit{Medicago truncatula} & 10,000 & 102,580 & 467 & 0.50 \\ 
   \hline
\end{tabular}
\caption{
    Descriptive statistics for each dataset used,
    with the chosen window sizes.
    See text for how window sizes were chosen;
    in each window we performed PCA,
    computed the percent varince explained by the top two principal components;
    the mean across windows is given in the last column.
    \label{tab:data_stats}
}
\end{table}

@@

\plr{I'm talking about flipping signs up here, even though in the code we do it when computing variances. Do you think it makes sense here?}
Since eigenvectors are still only defined up to sign,
when comparing between windows we choose the sign to best match each other:
after choosing $\pcone_1$, for instance, and $u$ is the first eigenvector obtained from the covariance matrix
for window $j$,
then $\pcone_j = \pm u$,
where the sign is chosen according to which of 
$\sum_i (\pcone_{i1} - u_i)$ or
$\sum_i (\pcone_{i1} + u_i)$ 
is smaller.

@@

The choice of window length entails a balance between several factors.
In very short windows, genealogies of the samples will only be represented by a few trees,
so variation between windows represents demographic noise rather than meaningful variation in population structure.
Longer windows generally have more distinct trees (and SNPs), 
allowing for less noisy estimation of local population structure.
However, to better resolve meaningful signal, i.e., differences in population structure along the genome, 
we would like reasonably short windows.
Window length choice therefore entails a signal versus noise tradeoff
in the estimates of population structure.
% If we use the first principal component as a measure of population structure, 
% then to choose the best window length,
% we need to find a balance between the standard error of the first principal components for each window 
% and the standard deviation between windows. 
Since we summarize population structure using relative positions in the principal component maps,
we quantify ``noise'' as the standard error of a sample's position on PC1 in a particular window,
averaged across windows and samples,
and ``signal'' as the standard deviation of the sample's position on PC1 over all windows,
averaged over samples.
(Recall that the signs for PCs are chosen to match each other.)
Then, the mean variance across windows is
\begin{align*}
    \sigma_\text{signal}^2
    = 
    \frac{1}{N} \sum_{j=1}^{N}
        \frac{1}{L}\sum_{i=1}^{L}\left ( \pcone_{ij} -\overline\pcone_{j} \right )^{2} ,
\end{align*}
where $\overline\pcone_j = (1/N) \sum_{j=1}^N \pcone_{ij}$.
We estimate the standard error for each $\pcone_{ij}$ using the block jackknife \citep{efron1982jackknife}:
we divide the $j^\text{th}$ block into 10 equal-sized pieces,
and let $\pcone_{ij,k}$ denote the first principal component of this region found after removing the $k^\text{th}$ piece;
then the estimate of the squared standard error is
$\sigma^2_{ij} = \frac{9}{10} \sum_{k=1}^{10} ( \pcone_{ij,k} - \frac{1}{10} \sum_{\ell=1}^{10} \pcone_{ij,\ell} )^2$,
and we measure noise by averaging over samples and windows:
\begin{align*}
    \sigma^2_\text{noise}
    &=
    \frac{1}{N} \sum_{j=1}^{N} \frac{1}{L}\sum_{i=1}^{L} \sigma^2_{ij} .
\end{align*}

We calculated $\sigma^2_\text{signal}$ and $\sigma^2_\text{noise}$
for a range of window sizes,
and for our main results
chose the smallest window for which $\sigma^2_\text{signal}$ was consistently larger than $\sigma^2_\text{noise}$;
the values for various window sizes across \textit{Drosophila} chromosomes are shown in Table \ref{tab:window_sizes},
and the choices for all taxa are in Table \ref{tab:data_stats}.
In practice, even substantially different window sizes result in the same genome-wide patterns.
We also chose windows to each consist of the same number of neighboring SNPs.
We compared to the results when windows are each of the same physical length (in bp) but of the same average length,
and results were nearly identical.

% # Variances within and between windows for Drosophila, multiplied by 1000:
% w <- structure(list(chrom = structure(c(1L, 1L, 2L, 2L, 3L, 3L, 4L, 
% 4L, 5L, 5L), .Label = c("2L", "2R", "3L", "3R", "X"), class = "factor"), 
%     type = structure(c(1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 2L
%     ), .Label = c("noise", "signal"), class = "factor"), `100` = c(2.05209, 
%     2.75625, 2.18089, 2.77729, 2.07936, 2.601, 1.95364, 2.58064, 
%     2.48004, 2.61121), `500` = c(1.64025, 2.69361, 1.91844, 2.704, 
%     1.99809, 2.52004, 1.764, 2.51001, 2.04304, 2.43049), `1000` = c(1.18336, 
%     2.22784, 1.63216, 2.65225, 1.64025, 2.401, 1.43641, 2.44036, 
%     1.53664, 2.304), `10000` = c(0.169, 0.676, 0.576, 2.31361, 
%     0.73441, 1.681, 0.58564, 1.96249, 1.61604, 0.324), `1e+05` = c(0.03969, 
%     0.31329, 0.13456, 1.82329, 0.24649, 1.89225, 0.20164, 1.39876, 
%     0.169, 1.14244)), .Names = c("chrom", "type", "100", "500", 
% "1000", "10000", "1e+05"), row.names = c(NA, -10L), class = "data.frame")
% print(xtable(w),include.rownames=FALSE)

\begin{table}[ht]
\centering
    \begin{tabular}{cccrrrrr}
  \hline
        & & & \multicolumn{5}{c}{window length (SNPs)} \\
 & chrom.\ arm  & & 100 & 500 & 1,000 & 10,000 & 100,000 \\ 
  \hline
    & 2L & $\sigma^2_\text{noise}$  & 2.05  &  1.64  &  1.18  &  0.17  &  0.04 \\
    & 	 & $\sigma^2_\text{signal}$ & 2.76  &  2.69  &  2.23  &  0.68  &  0.31 \\
    & 2R & $\sigma^2_\text{noise}$  & 2.18  &  1.92  &  1.63  &  0.58  &  0.13 \\
    & 	 & $\sigma^2_\text{signal}$ & 2.78  &  2.70  &  2.65  &  2.31  &  1.82 \\
    & 3L & $\sigma^2_\text{noise}$  & 2.08  &  2.00  &  1.64  &  0.73  &  0.25 \\
    & 	 & $\sigma^2_\text{signal}$ & 2.60  &  2.52  &  2.40  &  1.68  &  1.89 \\
    & 3R & $\sigma^2_\text{noise}$  & 1.95  &  1.76  &  1.44  &  0.59  &  0.20 \\
    & 	 & $\sigma^2_\text{signal}$ & 2.58  &  2.51  &  2.44  &  1.96  &  1.40 \\
    & X  & $\sigma^2_\text{noise}$  & 2.48  &  2.04  &  1.54  &  1.62  &  0.17 \\
    & 	 & $\sigma^2_\text{signal}$ & 2.61  &  2.43  &  2.30  &  0.32  &  1.14 \\
   \hline
\end{tabular}
\caption{
    Measures of signal and noise,
    computed separately for each chromosome arm in the \textit{Drosophila} dataset,
    at different window sizes.
    All values are multiplied by $1,000$
    (so typical variation is of order of 50\% of the actual values).
    Starting at windows of 1,000 SNPs, the signal (variation of PC1 between windows)
    starts to be substantially larger than the noise (standard error of PC1 for each window).
} \label{tab:window_sizes}
\end{table}

\subsection{Similarity of population structure between windows}

We think of local population structure as being summarized by \emph{relative} position of the sampmles
in the space defined by the top principal components.
Since rotations of this space imply identical population structure,
we do not compare population structure of different genomic windows using the PCs directly, 
but instead compare the low-dimensional approximations of the local covariance matrices
obtained using the top $k$ PCs.
(For results shown here, we use $k=2$;
results using larger numbers of PCs were nearly identical.)
We do this, rather than using the entire covariance matrix, 
both for computational efficiency 
and because the top PCs summarize important population structure, 
so using only these should reduce the effect of noise.
To remove the effect of variation in mutation rate between windows,
we also rescale each covariance matrix so that the underlying data matrix has trace norm equal to one.
To do this, define the $N \times k$ matrix $V(i)$ so that the $\ell^\text{th}$ column of $V(i)$
is equal to the $\ell^\text{th}$ princpal component of the $i^\text{th}$ window,
multiplied by $\sqrt{ \lambda_{\ell i} / \sum_{m=1}^k \lambda_{m i} }$.
% \begin{equation}
%     \begin{aligned}
%         V_{i1}&=\sqrt{\frac{\lambda _{1i}}{\lambda _{1i}+\lambda _{2i}}}PC1_{i} ,
%         \qquad
%         V_{i2}&=\sqrt{\frac{\lambda _{2i}}{\lambda _{1i}+\lambda _{2i}}}PC2_{i} 
%     \end{aligned}
% \end{equation}
Then, the rescaled, $k$-dimensional covariance matrix for the $i^\text{th}$ window is
\begin{align}
    M(i) &= \sum_{\ell=1}^k V_{i\ell} V_{i\ell}^T .
\end{align}
% \begin{align}
%     M_{i} &= \frac{\lambda_{1i}PC1_{i}PC1_{i}^{T}+\lambda_{2i}PC2_{i}PC2_{i}^{T}}{\lambda_{1i}+\lambda_{2i}} 
% \end{align}

We then use
Euclidean distance $D_{ij}$ between the matrices $M(i)$ and $M(j)$ 
to measure the similarity of population structure for the $i^\text{th}$ window and $j^\text{th}$ window. 
Since 
$\sum_{ij} (A_{ij}-B_{ij})^2 = \sum_{ij} (A^2_{ij} + B^2_{ij}) - 2 \tr(A^T B)$,
and $M(i) = V(i) V(i)^T$,
then due to the orthogonality of eigenvectors and the cyclic invariance of trace,
$D_{ij}$ can be computed efficiently as
\begin{align}
    D_{ij} 
    = 
    \frac{ \sum_{\ell=1}^k \lambda_{\ell i}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell i})^2 }
    + \frac{ \sum_{\ell=1}^k \lambda_{\ell j}^2 }{ (\sum_{\ell=1}^k \lambda_{\ell j})^2 }
    - 2 \sum_{\ell, m=1}^k (V(i)^T V(j))^2_{\ell m} .
\end{align}
% from the code for pc_dist():
% #' In the unweighted case, suppose that (u) and (v) are sets of vectors, and that 
% #'    A = a_1 * u_1 u_1^T + ... + a_j * u_j u_j^T = U diag(a) U^T
% #'    B = b_1 * v_1 v_1^T + ... + b_j * v_k v_k^T = V diag(b) V^T,
% #' where U is the matrix whose columns are (u), and likewise for V.
% #' Then 
% #'    (A-B)^T (A-B) = A^T A + B^T B - A^T B - B^T A
% #' and so
% #'    ||A-B||^2 = ||A||^2 + ||B||^2 - 2 tr( A^T B )
% #' By the cyclic invariance of trace, if X = U^T V then
% #'    tr( A^T B ) = tr( U diag(a) U^T V diag(b) V^T ) = tr( diag(a) X diag(b) X^T ) .
% \begin{align}
%     \begin{split}
%     D_{ij} &= 
%         \left \{
%                 \left ( 
%                     V_{i1}\cdot V_{i1}^{} 
%                 \right )^{2}
%                 + \left ( V_{i2}\cdot V_{i2}^{} \right )^{2}
%                 + \left ( V_{j1}\cdot V_{j1}^{} \right )^{2} 
%                 +\left ( V_{j2}\cdot V_{j2}^{} \right )^{2} 
%         \right. \\ 
%         & \left. \qquad {}  
%             -2\left [ 
%                 \left ( V_{i1}\cdot V_{j1}^{} \right )^{2}
%                 + \left ( V_{i1}\cdot V_{j2}^{} \right )^{2}
%                 + \left ( V_{i2}\cdot V_{j1}^{} \right )^{2} 
%                 + \left ( V_{i2}\cdot V_{j2}^{} \right )^{2}
%             \right ]
%         \right \}^{1/2}
%     \end{split}
% \end{align}

\subsection{Visualization of results}

We use multidimensional scaling (MDS) to visualize relationships between windows
as summarized by the dissimilarity matrix $D$,
that produces a set of $m$ coordinates for each window that
give the arrangement in $m$-dimensional space that best recapitulates the original distance matrix.
For results here, we use $m=2$ to produce one- or two-dimensional visualizations of relationships between windows' population structure.

We then locate variation in population structure along the genome
by choosing collections of windows that are nearby in MDS cooridinates,
and map their positions along the genome.
Their common population structure is identified by extracting the corresponding genomic regions
and performing PCA on all, aggregated regions.

\subsection{Software}

The methods described here
are implemented in an open-source R package
available at \url{https://github.com/petrelharp/lostruct},
as well as scripts to perform all analyses from VCF files
at various parameter settings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

In all these 3 species, PCA plots vary along the genome in a systematic way, showing strong chromosome-scale correlations.
This implies that variation is truly due to population structure, 
since fluctuations due to demographic noise are not expected to show long distance correlations. 
Below, we display the results and investigate likely underlying causes.
